Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-18 ms/review-responses.tex

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts > 2bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
The mean of the IBS tract distribution is calculated by building a (very large) vector of lengths of all IBS tracts between all pairs of individuals and taking the mean over this entire vector. This distribution is quite strongly skewed (i.e. see figure 4 and the skew(IBs) panel of fig S1) such that the highest density is in low values. We explain the depressed mean value by noting that at low neighborhood sizes the coalescence time of distant (in space) individuals is inflated, which allows more time for recombination and mutation to break up IBS tracts. The flip side of this is that there are then more IBS tracts in total across a random sample of individuals, explaining the corresponding inflation in the number of tracts. We also see an increase in the number of very long IBS tracts at low neighborhood sizes, which corresponds to an excess of very recent coalescent events for groups of nearby individuals. 
\cjb{need to think of something to say we did about this in the text}

}

\begin{point}{}
    The authors use K to denote both carrying capacity \llname{ll:K_defn}
    and population density \llname{ll:K_again}. 
    It might be better to use a different notation for these quantities since carrying capacity is fixed while density is an emergent quantity in the non-Wright-Fisher model. Use of K to denote carrying capacity and density is a bit confusing. For example, on \llname{ll:K_more} it is said that ' the ``population density'' (K) and ``mean lifetime'' (L) parameters were the same in all simulations'. Here K seems to indicate carrying capacity rather than density? The latter is an emergent quantity and varies across simulation runs?
\end{point}

@@

\reply{
    We agree that this distinction is worth emphasizing!
    We've adjusted our language to hopefully remind the reader that
    $K$ is a parameter that controls population density, rather than being equal to it,
    at \llname{ll:K_again} and \revref{} and \llname{ll:K_more}.
}

\begin{point}{}
    Concerning the non-Wright-Fisher model used, it would be helpful to emphasize that some of the parameters are emergent in contrast to Wright-Fisher model. For example, on Page 11, lines 306-308, the author's goal was to look at census size variation and variation in other quantities. This would be better understood if to emphasize that these parameters are emergent properties in the non-Wright-Fisher model used.
\end{point}

\reply{
    We have added to the text at the beginning of the results to emphasize that this analysis is necessary because these parameters are emergent rather than fixed \llname{emergentparams}. 
}

\begin{point}{\revref}
    Perhaps 'Demographic Inference' might better reflect the content of this section.
\end{point}

\reply{
Good suggestion -- we have changed the section heading to ``demographic inference'' \revref.
}

@@

\reply{
We have revised this sentence to clarify that we aim to produce a variation in mean phenotype of two standard deviations across the landscape \llname{phenotypescalingsd}.
}

\begin{point}{\revref}
    In the sentence, 'We also examined p values for systemic inflation' I think the authors meant 'systematic inflation'.
\end{point}

\reply{
    Whoops; thanks. Fixed.
}

\begin{point}{}
    Please correct the legend in \autoref{fig:genparams}:
    must be `spatial model' and `random mating' model.
\end{point}

\reply{
Thanks for catching our confusing legend title placement!
We have moved ``model'' to after ``spatial'' as suggested.
}

\begin{point}{}
    Optional: a dashed line in \autoref{fig:genparams}
    that shows the total carrying capacity of $50 \times 50 \times 5=12500$ would be helpful.
\end{point}

@@

\reply{
    This is a good suggestion, but we decided to not include this as we don't have straightforward expectations for the other parameters shown. 
}

\begin{point}{\revref}
    The phrase `affect summaries of variation' is better to replace with `summaries of genetic variation'.
\end{point}

@@

\begin{point}{}
    The github link to the code didn't work for me. I assume it will be made public later, but at this point I can't tell whether the code is available/useable.
\end{point}

\reply{
Apologies, I had forgotten to make the repo public. The link should work for everyone now. 
}

%%%%%%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
    The present study deals with a ``hot topic'' in spatial population genetics. Most
    inferential and descriptive methods in statistical spatial population
    genetic rely on a discrete approximation of space and it is not clear what
    impact this approximation may have when individuals migrate along a
    continuum instead. Spatial patterns in sampling is also another major issue
    which is often simply dismissed, mainly because of the paucity of
    statistical methods to deal with it. This work touches on these important
    issues in a timely manner.
\end{quote}

\begin{quote}
    Although I was enthusiastic about the topic, I was quite disappointed with
    the core of the study, i.e., the forward-in-time simulation of populations in
    continuous space. The field has been struggling with this issue for decades --
    examples of spectacular failures like the Wright-Malecot model (see
    Felsenstein's ``pain in the torus'' article, 1975) or, more recently, the
    ``mugration'' or ``discrete trait analysis'' model in phylodynamics (see De Maio
    et al. 2015) have probably mostly harmed our research field -- that one cannot
    make the economy of using a sound probabilistic model for generating
    geo-referenced genetic data. It does not seem to be the case here
    unfortunately.  
\end{quote}

@@

\reply{
    This is a good point also raised by other reviewers. We have added an Appendix comparing our model to a stepping-stone, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models. See Appendix 1. We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases. 
}

\begin{point}{}
    Figure 2: I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From lines 166->171, it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

\reply{
    Census population in the random mating model scales with ``neighborhood size'' because of two effects -- the edge scaling and the competition

@@

    \cjb{need a better answer here...}
}

\begin{point}{}
    Line 729-731: ``Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]''. Do the authors think that the spatial processes would still play a minor role when neighborhood sizes exceed 100-1000 if the habitat was larger than that taken in the present simulations? It would also probably be useful to mention that neighborhood sizes given in Table 1 should be compared with extreme caution since the size of the corresponding habitats vary across species. More generally, I suspect that the size of the habitat has a substantial impact on the vast majority of statistics examined in this study. Indeed, the mean parent-offspring distance, which is at the core of the definition of Wright's neighborhood size, is only small or large relative to the size of the habitat.
\end{point}

\reply{
This is a good point. Wright's work \citep{Wright1943} suggests some aspects of genetic variation such as variance in allele frequencies and inbreeding coefficients can be estimated by looking only at what he would later \citep{Wright1946} call ``neighborhood size'', but certainly other aspects like the number of segregating sits will also depend on total landscape size. We now note on \llname{popsizecaveat} that we have evaluated only one landscape size, and have added a sentence to the discussion noting that exploration of these patterns in varying landscape sizes is an important avenue for further research \llname{landscapesizecaveat}.

@@

\reply{
Apologies, it was accidentally set to private. The link should work now. 
}

\begin{point}{}
    Definitions and interpretations. There are quite a large number of metrics discussed in Figure 3B, and it's a lot to take in. It might be helpful to have a table with a reminder of what the metric is, its interpretation, and how it is computed.
\end{point}

\reply{
    We have included a table describing the summary statistics in Figure \ref{table:sumstats}.
}

\begin{point}{}
    Notation: ``Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates. Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$.'' I think x may be overloaded here, or I'm confused. Clarify?
\end{point}

@@

\reply{
\cjb{Peter and Andy -- ideas here?}
}

\begin{point}{}
    When introducing the ``spatial model'' as opposed to this ``random model,'' the more concrete illustration in Figure 1 hasn't yet been referenced, which makes it harder to follow. It would be helpful to introduce this figure with the model. Additionally, when Figure 1 is introduced, the order is from right to left (random, then point, then midpoint). It would be helpful to rearrange the figure to mirror what's in the text.
\end{point}

@@

\reply{
We have rearranged the figure as suggested.

}

\begin{point}{}
    Not sure I follow this example: ``Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5.''
\end{point}

@@

\reply{
}

\begin{point}{\revreffull{1}{11}}
    Minor typo (through vs though): \emph{This occurs because, even through the ``population density'' (K) and ``mean lifetime'' (L) parameters...}
\end{point}

@@

\reply{
Thanks, this sentence has been revised and fixed. 
}

\begin{point}{}
    Define NS abbreviation in Figure \ref{fig:dafdists}.
\end{point}

### 2019-11-18 ms/spaceness.tex

%%%%%%%%%%%
\subsection{Demographic Inference}

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics \llabel{spatialgenetic}.

@@

We chose our demographic model so that every individual has on average $1/L$ offspring
each time step,
and if the local population density of an individual is $n$,
then their probability of survival until the next time step is
(equation \eqref{eqn:p_survival}):
\begin{equation}
    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
\end{equation}
We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
in sparsely populated areas --
otherwise, an isolated individual might live for a very long time.
Since $1 - p \approx n / (K (1 + L))$,
mortalitity goes up roughly linearly with the number of neighbors (on a scale given by $K$),
as would be obtained if, for instance, mortality is due to agonistic interactions.
Ignoring in/outmigration,
a region is at demographic equilibrium if the per-capita probability of death
is equal to the birth rate,
i.e., if $1 - p = 1/L$.
(Note that there is no effect of age in the model,
which would make the analysis more complicated.)
Solving this for $n$,
we get that in a well-mixed population, the equilibrium density should be around
\begin{equation}
    n &= K \frac{L+1}{L-1}
\end{equation}
individuals per unit area.
At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
However, in practice this model leads to a total population size which is around $K$
multiplied by total geographic area
(but which depends on $\sigma$, as discussed above).
The main reason for this is that
since offspring tend to be near their parents, individuals tend to be ``clumped'',
and so experience a higher average density than the ``density'' one would compute
by dividing census size by geographic area \citep{lloyd1967crowding}.
To maintain a constant expected total population size
would require making (say) $K$ depend on $\sigma$;
however, typical local population densities might then be more dissimilar.

>>>>

%%%%%%%%%%%
\subsection{Demographic Inference}
\revpoint{1}{4}

@@

Even though certain aspects of population demography depend on the scale of spatial interactions, it still could be that population genetic variation is well-described by a well-mixed population model. Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}. However, the behavior of most common population genetic summary statistics other than Tajima's $D$ \citep{Stadler2009} has not yet been described in realistic geographic models. Moreover, as we will show, spatial sampling strategies can affect summaries of genetic variation at least as strongly as the underlying population dynamics. \revpoint{1}{9}

@@

We chose our demographic model so that every individual has on average $1/L$ offspring
each time step,
and if the local population density of an individual is $n$,
then their probability of survival until the next time step is
(equation \eqref{eqn:p_survival}):
\begin{equation}
    p = \min\left( 0.95, \frac{1}{1 + n / (K (1+L))} \right) .
\end{equation}
We capped survival at 0.95 so that we would not have exceptionally long-lived individuals
in sparsely populated areas --
otherwise, an isolated individual might live for a very long time.
Since $1 - p \approx n / (K (1 + L))$,
mortalitity goes up roughly linearly with the number of neighbors (on a scale given by $K$),
as would be obtained if, for instance, mortality is due to agonistic interactions.
Ignoring in/outmigration,
a region is at demographic equilibrium if the per-capita probability of death
is equal to the birth rate,
i.e., if $1 - p = 1/L$.
(Note that there is no effect of age in the model,
which would make the analysis more complicated.)
Solving this for $n$,
we get that in a well-mixed population, the equilibrium density should be around
\begin{equation}
    n = K \frac{L+1}{L-1}
\end{equation}
individuals per unit area.
At this density, the per-capita death rate is $1/L$, so the mean lifetime is $L$.
This equilibrium density is \emph{not} $K$, but (since $L=4$) is two-thirds larger.
However, in practice this model leads to a total population size which is around $K$
multiplied by total geographic area
(but which depends on $\sigma$, as discussed above).
The main reason for this is that
since offspring tend to be near their parents, individuals tend to be ``clumped'',
and so experience a higher average density than the ``density'' one would compute
by dividing census size by geographic area \citep{lloyd1967crowding}.
To maintain a constant expected total population size
would require making (say) $K$ depend on $\sigma$;
however, typical local population densities might then be more dissimilar.
